Synthetic algebraic geometry and the phenomenon of nongeometric sequents

### 2018-04-06 slides-pssl103.tex
Some reordering

\subsection{Synthetic differential geometry}

@@

\subsection{Toposes}

\begin{frame}{The internal universe of a topos}
  A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier, for instance
  \begin{itemize}
    \item \hil{$\boldsymbol{\Set}$}, the category of sets;
    \item \hil{$\boldsymbol{\Sh(X)}$}, the category of set-valued sheaves over
    a space~$X$;
    \item \hil{$\boldsymbol{\Eff}$}, the ef{}fective topos (roughly: a
    category of data types).
  \end{itemize}

      \item \intex{
        $\Eff \not\models \text{``Any Turing machine halts or doesn't halt.''}$
      }{\textcolor{red!80}{\xmark}}{
        External meaning: There is a \hil{halting oracle} which determines whether
        any given machine halts or doesn't halt.
      }

      \item \intex{
        $\Sh(X) \not\models \text{``Any cont. function with opposite signs has
        a zero.''}$
      }{\textcolor{red!80}{\xmark}}{
        External meaning: Zeros can locally be picked \hil{continuously} in
        continuous families of continuous functions.
      }
    \end{itemize}
  \end{changemargin}

@@

\begin{frame}{Approaches to algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip
  \pause

@@

@@

\begin{frame}{Properties of the affine line}
  \begin{itemize}
    \item $\affl$ is a local ring:
    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
    \text{$y$ \inv} \]
    \item $\affl$ is a field:
    \begin{align*}
      \neg(x = 0) &\Longleftrightarrow \text{$x$ \inv} \\
      \neg(\text{$x$ \inv}) &\Longleftrightarrow \text{$x$ nilpotent}
    \end{align*}
    \item $\affl$ satisfies the axiom of microaffinity.
    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
    does \emph{not not} have a zero.
    \item $\affl$ is of unbounded Krull dimension.
    \item Any function $\affl \to \affl$ is a polynomial.
  \end{itemize}
\end{frame}

>>>>

\subsection{Approaches to algebraic geometry}

\begin{frame}{Approaches to algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip

\subsection{Toposes as universes}

\begin{frame}{Toposes as mathematical universes}
  A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier, for instance
  \begin{itemize}
    \item \hil{$\boldsymbol{\Set}$}, the category of sets;
    \item \hil{$\boldsymbol{\Sh(X)}$}, the category of set-valued sheaves over
    a space~$X$;
    \item \hil{$\boldsymbol{\Eff}$}, the ef{}fective topos (roughly: a
    category of data types).
  \end{itemize}

      \item \intex{
        $\Eff \models \text{``Any Turing machine halts or doesn't halt.''}$
      }{\textcolor{red!80}{\xmark}}{
        External meaning: There is a \hil{halting oracle} which determines whether
        any given machine halts or doesn't halt.
      }

      \item \intex{
        $\Sh(X) \models \text{``Any cont. function with opposite signs has
        a zero.''}$
      }{\textcolor{red!80}{\xmark}}{
        External meaning: Zeros can locally be picked \hil{continuously} in
        continuous families of continuous functions.
      }
    \end{itemize}
  \end{changemargin}

@@

@@

\subsection{The big Zariski topos}

@@

\subsection{Synthetic constructions}

@@

\subsection[Quasicoherence]{Synthetic quasicoherence}

\begin{frame}{Properties of the affine line}
  \begin{itemize}
    \item $\affl$ is a local ring:
    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
    \text{$y$ \inv} \]
    \item $\affl$ is a field:
    \begin{align*}
      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible} \\
      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
    \end{align*}
    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
    \varepsilon^2 = 0 \}$.
    \item Any function $\affl \to \affl$ is a polynomial.
    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
    does \emph{not not} have a zero.
    \item $\affl$ is of unbounded Krull dimension.
  \end{itemize}
\end{frame}
